Experience in Modeling a Microcontroller Instruction Set Using B

### 2012-05-21 trunk/papers/ArticleZ80/source/paper.tex

This work presents a formal model of the instruction set of the Z80
microcontroller~\cite{Z80_manual}. This microcontroller was chosen
because of several factors: it contains essential and common concepts
of microcontrollers and even microprocessors; it has extensive
available documentation; and it has been widely used.  A first,
partial, Z80 model presented in \cite{Valerio_SBMF09} was fully
verified, but it used constructs that made it impossible to animate.

@@

\item[Power] contains the standard definition of exponentiation, as
  well as some basic properties, e.g. values of different powers of
  two. Given this definition, these properties have been proved
  interactively with the Atelier-B prover.

\item[Hardware Library] defines bit vectors with size 8 and 16, basic
  functions to manipulate bit vectors and important
  assertions. Details are given in sections
  \ref{subsec:HardwareLibrary1}, \ref{subsec:HardwareLibrary2} and
  \ref{subsec:HardwareLibrary3}.

@@

We model bits as a set of integers: $\textit{BIT} =
\textit{0..1}$. The negation is an unary function on bits and it is
defined below. The module also provides lemmas on negation that may be useful for the
users of the library to develop proofs.

@@

Similarly, several other general functions and lemmas were created for all other data types.
Some times, we need to create arithmetic and logic functions that are more specific for a
determined microprocessor, like Z80 that is presented in the section \ref{sec:z80}.

The definitions, functions and properties from hardware library and types library are difficult to verify, otherwise these libraries can be reused in several others microprocessors without to verify again. Nearly 50\% of all these proof obligations shown in the Table~\ref{tab:types} are verified automatically or with few user pass\footnote{The user pass allows to filter proof tactics that aid the prover to find the proof, this is a feature provided by Atelier B.}. But many others proof obligations are difficult to verify because they involve functions that contain complex arithmetic expressions to manipulate with the interactive prover of AtelierB. However, all these 531 proof obligations were verified completely.

@@

To avoid such inconsistencies, we state separately the required
properties in the \textit{ASSERTIONS\/} clause: this causes
corresponding proof obligations to be generated and verified. However,
we have not been able to build the corresponding proofs using the 
rule base included in the free distribution of AtelierB. So we
formalized new proof rules to check these properties in AtelierB and
used them in the interactive prover of AtelierB.

@@

This section shows the elements that make up the different types of
instructions. The main elements are specified in the microcontroller module Z80
and parts of it are presented below.  Basically, the state of microcontroller is
formed by data from program counter, ports, registers, etc. The transitions
between the states are actioned by execution of instruction or an external
action. Groups of registers are represented by the variables of the
specification states that appear in clause \textit{VARIABLES}. The declaration
of valid states of variables is in the \textit{INVARIANT} clause and the initial
state is defined in the \textit{INITIALISATION} clause. The assembly
instructions are defined by the clause \textit{OPERATIONS} and, in general, each
instruction has three actions: update the program counter, update the flag
register and its main effect.
% Several functions were created to Z80.
General functions that can be used in other microcontroller models are defined
in the modules of data types. Specific functions of the microcontroller are
defined in the ALU (arithmetic logic unit) module.The main module includes an instance of the memory module and accesses the
definitions from basic data types modules and the \textit{ALU} module.

@@

The Z80 CPU includes alternative set of accumulator, flag and general registers.
The CPU contains a stack pointer ($\textit{sp}$), program counter
($\textit{pc}$), two index registers ($\textit{ix}$ and $\textit{iy}$), an
interrupt register ($\textit{i\_}$), a refresh register ($\textit{r\_}$), two
bits ($\textit{iff1}$, $\textit{iff2}$) used to control the interruptions, a
pair of bits to define the interruption mode ($\textit{im}$) and the input and
output ports ($\textit{i\_o\_ports}$). These definitions are represented by
\textit{INVARIANT}.

@@

% Features of microcontroller
The internal registers contains 176 bits of read/write memory that are
represented by identifiers used as parameters in the instructions . It includes
two sets of six general purpose registers which may be used individually as
8-bit registers or as 16-bit register pairs.  The working registers are
represented by variable $\textit{rgs8}$. The domain of $\textit{rgs8}$
($\textit{id\_reg8}$) is a set formed by identifiers of registers of 8 bits.
These registers are called main register set (\textit{a0,b0,c0,d0,e0,f0,h0,l0})
and alternate register set (\textit{a\_0,b\_0,c\_0,d\_0,e\_0,f\_0,h\_0,l\_0}).
These registers can be accessed in pairs, forming 16-bits, resulting in another
set of identifiers of 16-bits registers, named $\textit{id\_reg16}$. These sets
are represented below.

@@

\textbf{Assuring the absence of overflow:}
 \emph{To assure that an overflow does not happen, the developer can add in the invariant this 
 expression:\\(\textit{bv\_get(rgs8(f0),0)} $\neq$ 1 $\land$
 \textit{bv\_get(rgs8(f0),2) $\neq$ 1}).
% When a overflow happen,it can be dangerous. So the developer may restrict its use. However, this restriction can also
% become more difficult to verify the model.
}

@@

There is a specific function to refresh the flag register, it is named $\textit{update\_reg}$ $\textit{\_flag}$. It is typed and defined as follow:\it update\_flag\_reg \rm $\in$ \rm (\it BIT  $\times$  \it BIT  $\times$  \it BIT  $\times$  \it
BIT $\times$  \it BIT  $\times$  \it BIT\rm) $\fun$  \rm (\rm \{\it f0\rm \}  $\times$  \it BYTE\rm ).  This function is defined as:\\

@@

In general, the instructions can access all memory addresses, but it is dangerous. The user may mistakenly
access and change data in memory. For added  security, it is important that the program instructions has
limited access by region. Thus the designer can specify address regions in a refinement model to restrict the access from
instructions. The address regions can be specified using constants
($\textit{PROGRAM\_R\_ADR,DATA\_R\_ADR,STACK\_R\_ADR}$), these define respectively a range of restricted
address for: programs instructions, general purpose data and data stack.
%
%$
%\begin{array}{l}
%\textit{PROGRAM\_R\_ADR} = 0..16384 \land\\
%\textit{DATA\_R\_ADR} = 16385..49151 \land\\
%\textit{STACK\_R\_ADR} = 49152..65535
%\end{array}
%$

@@

\textbf{Preserving the memory consistency:} \emph{In general, the access to
some address regions is dangerous. Then, each instruction has a specific pre-condition that verifies
if the new address memory, that will be updated, is a member of its region. For example, the $\textit{PUSH}$
program instruction permits write only in the region of stack ($\textit{STACK\_R\_ADR}$\rm).}

@@

There are many functions in the module \textit{ALU}. In general, the definition of these 
functions use basic definitions or previously defined functions. For example, the function
$\textit{half8UCHAR}$ is used to get the half part of $\textit{UCHAR}$ value.
It is important to know the half carry and it is used in the function $\textit{add8UCHAR}$. \\

@@

A related function to subtract operation is $\textit{subtract8UCHAR}$. There are the same functions for the
$\textit{SCHAR}$ type, they are respectively $\textit{add8SCHAR}$ and $\textit{subtract8SCHAR}$, all these
functions are of 8 bits ($\textit{BYTE}$) and defined similarly. In the same way, the arithmetic functions for 16
bits ($\textit{BV16}$) are defined.
The module ALU has several others functions, for example:%but only some simplest functions also are explained below:
%\begin{itemize}
  %\item \it inc  $\in$ \it BYTE $\fun$  \it BYTE \rm - It receives a byte and
  %returns its increment. There is a similar function named $\textit{dec}$.
%  \item 
\it instruction\_next  $\in$  USHORT  $\fun$  USHORT \rm - It receives the  
  actual value from program counter register (\textit{pc}) and returns its increment;
% \item 
and \it is\_negative  $\in$  \it BYTE  $\fun$  \it BIT \rm - It returns the most significant bit,
 in other words, the signal bit.
  % \item \it is\_zero  $\in$  \it BYTE  $\fun$  \it BIT \rm - It returns 1 if the received
 % byte is zero, otherwise returns 0.
  %\item \it update\_refresh\_reg \rm - It receives a byte and returns its increment until the seventh bit.
%\end{itemize}
As the logic functions that are defined in the \textit{BYTE} and \textit{BV16}
module are included in the \textit{ALU} module, they can be seen and used
directly in the \textit{ALU} and \textit{Z80} modules.

@@

The other external actions are related to interruptions. Interruptions allow
devices to suspend a routine from CPU and start another service routine.
This service routine can exchange data or signals between CPU and external
devices. When a routine is finished, then the CPU comes back to the last routine
that was interrupted. Modeling details about interruptions can be consulted in
public project's repository.

@@

The animation in ProB is not as fast and practical as a simulation
with a dedicated tool (e.g. \cite{Simulator_z80}). ProB is a
general-purpose animator of formal specification, and addresses
computationally more difficult problems. Anyway, it is capable of
animating, and also possibly model checking, a formal model of the
instruction set of the Z80, which was our initial motivation in
constructing that model. One may simulate step-by-step sequences of
Z80 instructions and visualize the visited states of ports, memory and
registers. It is a useful tool to debug by creating a trace of the
visited states, offering to analyse constants, functions and
expressions, and search possible violations.

The construction of the model with B generates a large number of proof
obligations and the verification is a time-consuming task. We describe
here some details on this effort and some solutions we have devised to
reduce the effort.  In B, the generated proof obligations allow to
verify the data satisfy invariant properties, check important system
properties and that the expressions are well-defined (WD)\footnote{An
  expression is called ``well-defined'' (or unambiguous) if its
  definition assigns it a unique interpretation or value.}.

@@

% Porque o modelo não foi verificado completamente ?
There is still a large number of open proof obligations, i.e. we have
not yet found their proof, or they are not valid and our model is not
correct. Each such proof obligation requires interaction with a proof
assistant. We see that for a project of that size, a large effort
needs to be dedicated to find those proofs. One interesting feature of
the proof assistant is, once a proof has been found, it can
automatically apply it to other proof obligations, as it is often the
case that the same proof script discharges several proof obligations.

% ->Quanto tempo foi gasto nessa verificação ?  Aproximadamente quanto tempo de
% processamento elas reduziram ?
We had a previous experience with a similar effot, where we were able
to discharge all the proof obligations: that of the model of the Z80
instruction set designed without restricting ourselves to constructs
that may be animated with ProB, and without modeling external actions.
In that first experience, the verification took us about two months. In that second
project, we have been able to verify 3062 (e.g. roughly 80 \%) of 3631
proof obligations in one week.

In order to be able to verify such a large number of proof obligations
in such a short term, we employed two strategies: parallelism and
\emph{user pass}. AtelierB has an option to run any number of threads
in parallel and also to send proof obligations to other, possibly
remote, processes running AtelierB. With those options we had the
possibility to have eight concurrent instances of the prover. We thus
split the instruction set of the Z80 in eight sub-modules and checked
them concurrently (each instance of AtelierB may only verify one module
at a time).

@@

The Z80 formal model provides many benefits, because of the
verification of generated proof obligations guarantees: the correct
use of data types, the developed security properties and the
well-defined of all the expressions.  Furthermore, the designer has a
big flexibility to create new and specific security properties, this
is very
useful %TODO: COLOCAR APÓS TODOS OS BENEFICIOS PROPORCIONADOS COM A SIMULAÇÃO e COMPATIBILIDADE COM O PROB. (OBTER DO PARAGRAFO ANTERIOR DA SEÇAO VERIFICATION PROCESS).
to adjust the verification in accordance with the
requirements. Moreover, this example of model could complement the
existing documentation for users and assembly programmers.

But our goal is to use the model of the Z80 instruction set to extend
the scope of the B method up to the assembly
level~\cite{Dantas_SBMF08}. Indeed, we realized a case study where an
abstract B model had been refined up to the algorithmic model using
the classic development strategy. The algorithmic model was manually
compiled to a Z80 assembly program. We then constructed a B module
composed with the instructions of that program, and importing the
formal model of the Z80 instruction set. We then showed that this
assembly-level module was indeed a refinement of the original abstract
model.

@@

There exists several approaches to model hardware and instruction set
using B a.  The paper \cite{springerlink:Yuan2011} reports a method to
specify, design and construct sound and complete instruction set model
by stepwise refinement and formal proof using Event-B. It also
discusses desirable properties of an instruction set model. However,
our work \cite{Valerio_SBMF09} and \cite{Subotic2010} use the B
method, that seems quite appropriated to software development, because
it has an implementable language defined, called B0, a model-driven
approach \cite{Dantas_SBMF08} to develop software from the functional
specification level down to assembly, and tools to convert the models
to a programming language.

@@

% [Trabalhos futuros ] LLVM:Existe também a pretensão de especificar as
In the future, we plan to improve the support for the development of
software with the B method from functional specification to assembly
level, using the Z80 model presented in this work. For instance, the
mechanic compilation from B algorithmic constructs to assembly
platform is also envisioned. Finally, another important activity is to
develop a formal model of a platform used in LLVM
Compiler~\cite{DBLP:conf/cgo/LattnerA04}, which would enable us to
integrate different compilation techniques into our B-based framework.

>>>>

This work presents a formal model of the instruction set of the Z80
microcontroller~\cite{Z80_manual}. This microcontroller was chosen
because of several factors: it contains essential and common concepts
of microcontrollers and even microprocessors; it has extensive
available documentation; and it has been widely used.  A first,
partial, Z80 model presented in \cite{Valerio_SBMF09} was fully
verified, but it used constructs that made it impossible to animate 
with existing technology. %%DD

@@

\item[Power] contains the standard definition of exponentiation, as
  well as some basic lemmas, e.g. values of different powers of
  two, allproved interactively with the Atelier-B prover.
  This module is essential to establish the relationship between
  bit vectors and integer arithmetics. %%DD

\item[Hardware Library] defines bit vectors with size 8 and 16, basic
  functions to manipulate bit vectors and important lemmas. Details
  are given in sections \ref{subsec:HardwareLibrary1},
  \ref{subsec:HardwareLibrary2} and \ref{subsec:HardwareLibrary3}.

@@

We model bits as a set of integers: $\textit{BIT} =
\textit{0..1}$. The negation is an unary function on bits; its definition
follows. The module also provides lemmas on negation that may be
useful for the users of the library to develop proofs.

@@

Similarly, several other general functions and lemmas were created for
all other data types.  We sometimes need to create arithmetic and
logic functions that are more specific for a determined
architecture, as the Z80 presented in section \ref{sec:z80}.

%%DD As primeiras frases do parágrafo abaixo não faziam sentido!
The definitions, functions and properties from hardware library and
types library may be reused to model different instruction set
architectures. Nearly 50\% of all these proof obligations shown in
Table~\ref{tab:types} are verified automatically or with few user
pass\footnote{The user pass allows to filter proof tactics that aid
  the prover to find the proof, this is a feature provided by Atelier
  B.}. A large part of the proof obligations require long sessions
with the interactive provers, mostly because they contain complex
arithmetic expressions that the automatic provers in AtelierB have
difficulty handling. However, all these 531 proof obligations were
eventually verified.

@@

%DD forneci mais detalhes no paragrafo seguinte
To avoid such inconsistencies, we state separately the required
properties in the \textit{ASSERTIONS\/} clause: this changes the
status of these properties from axioms to lemmas and causes
corresponding proof obligations to be generated. However, we have not
been able to build the corresponding proofs using the rule base
included in the free distribution of AtelierB. In that situation, the
used to common approach to define new proof rules to check these
properties in AtelierB and to use them in the interactive prover of
AtelierB.

@@

This section shows the elements that make up the different types of
instructions. The main elements are specified in the microcontroller
module Z80 and parts of it are presented below.  Basically, the
microcontroller state is formed by the program counter, ports,
registers, etc. The stae transition occurs either as an instruction
execution or an external action. Groups of registers are represented
by the variables of the specification states that appear in clause
\textit{VARIABLES}. The declaration of valid states of variables is in
the \textit{INVARIANT} clause and the initial state is defined in the
\textit{INITIALISATION} clause. The assembly instructions are defined
by the clause \textit{OPERATIONS} and, in general, each instruction
has three actions: update the program counter, update the flag
register and its main effect.
% Several functions were created to Z80.
The functions specific to the microcontroller are defined in the ALU
(arithmetic logic unit) module. The main module includes an instance of
the memory module and accesses the definitions from basic data types
modules and the \textit{ALU} module.

@@

%%DD removi uma frase que não fazia sentido gramaticalmente.
The Z80 CPU contains a stack pointer ($\textit{sp}$), program counter
($\textit{pc}$), two index registers ($\textit{ix}$ and
$\textit{iy}$), an interrupt register ($\textit{i\_}$), a refresh
register ($\textit{r\_}$), two bits ($\textit{iff1}$, $\textit{iff2}$)
used to control the interruptions, a pair of bits to define the
interruption mode ($\textit{im}$) and the input and output ports
($\textit{i\_o\_ports}$). These definitions are represented by
\textit{INVARIANT}.

@@

% Features of microcontroller
The internal registers contains 176 bits of read/write memory that are
represented by identifiers used as parameters in the instructions. It
includes two sets of six general purpose registers which may be used
individually as 8-bit registers or as 16-bit register pairs.  The
working registers are represented by variable $\textit{rgs8}$. The
domain of $\textit{rgs8}$ ($\textit{id\_reg8}$) is a set formed by
identifiers of registers of 8 bits.  These registers are called main
register set (\textit{a0,b0,c0,d0,e0,f0,h0,l0}) and alternate register
set (\textit{a\_0,b\_0,c\_0,d\_0,e\_0,f\_0,h\_0,l\_0}).  These
registers can be accessed in pairs, forming 16-bits, resulting in
another set of identifiers of 16-bits registers, named
$\textit{id\_reg16}$. These sets are represented below.

@@

\textbf{Checking overflows:} \emph{To assure that an overflow does not
  happen, the developer can add in the invariant this
  expression:\\(\textit{bv\_get(rgs8(f0),0)} $\neq$ 1 $\land$
  \textit{bv\_get(rgs8(f0),2) $\neq$ 1}).  }

@@

There is a specific function to refresh the flag register, it is named
$\textit{update\_reg}$ $\textit{\_flag}$. It is defined as follow:\it
update\_flag\_reg \rm $\in$ \rm (\it BIT $\times$ \it BIT $\times$ \it
BIT $\times$ \it
BIT $\times$  \it BIT  $\times$  \it BIT\rm) $\fun$  \rm (\rm \{\it f0\rm \}  $\times$  \it BYTE\rm ).  This function is defined as:\\

@@

Instructions can access all the memory addresses and open the door to
the programmer mistakenly access and change data in memory. For added
security, it is desirable that the program instructions has limited
access by region. Thus the designer can specify address regions in a
refinement model to restrict the access from instructions. The address
regions can be specified using constants
($\textit{PROGRAM\_R\_ADR,DATA\_R\_ADR,STACK\_R\_ADR}$), these define
respectively a range of restricted address for: programs instructions,
general purpose data and data stack.
%
%$
%\begin{array}{l}
%\textit{PROGRAM\_R\_ADR} = 0..16384 \land\\
%\textit{DATA\_R\_ADR} = 16385..49151 \land\\
%\textit{STACK\_R\_ADR} = 49152..65535
%\end{array}
%$

@@

\textbf{Controlling memory access:} \emph{Each instruction has a
  specific pre-condition that verifies if the memory address that will
  be updated belongs. For example, the $\textit{PUSH}$ instruction
  permits write only in the stack area: $\textit{STACK\_R\_ADR}$.}

@@

There are many functions in the module \textit{ALU}. In general, the
definition of these functions use basic definitions or previously
defined functions. For example, the function $\textit{half8UCHAR}$ is
used to get the half part of $\textit{UCHAR}$ value. The value of the
half carry is important as it is used in the function $\textit{add8UCHAR}$. \\

@@

A related function is subtraction: $\textit{subtract8UCHAR}$. There
are also similar functions for the $\textit{SCHAR}$ type, they are
respectively $\textit{add8SCHAR}$ and $\textit{subtract8SCHAR}$, all
these functions are of 8 bits ($\textit{BYTE}$). In the same way, the
arithmetic functions for 16 bits ($\textit{BV16}$) are defined.  The
module ALU has several others functions, for
example:%but only some simplest functions also are explained below:
%\begin{itemize}
  %\item \it inc  $\in$ \it BYTE $\fun$  \it BYTE \rm - It receives a byte and
  %returns its increment. There is a similar function named $\textit{dec}$.
%  \item 
\it instruction\_next $\in$ USHORT $\fun$ USHORT \rm - It receives the
current value from program counter register (\textit{pc}) and returns
its increment;
% \item 
and \it is\_negative $\in$ \it BYTE $\fun$ \it BIT \rm - It returns
the most significant bit, in other words, the signal bit.
  % \item \it is\_zero  $\in$  \it BYTE  $\fun$  \it BIT \rm - It returns 1 if the received
 % byte is zero, otherwise returns 0.
  %\item \it update\_refresh\_reg \rm - It receives a byte and returns its increment until the seventh bit.
%\end{itemize}
As the logic functions that are defined in the \textit{BYTE} and
\textit{BV16} module are included in the \textit{ALU} module, they can
be seen and used directly in the \textit{ALU} and \textit{Z80}
modules.

@@

The other external actions are related to interrupts. Interrupts allow
devices to suspend a routine from CPU and start a service routine.
This routine can exchange data or signals between CPU and external
devices. When it finishes, then the CPU comes back to the interrupted
routine. Modeling details about interruptions can be consulted in
public project's repository.

@@

The animation in ProB is not as fast and practical as a simulation
with a dedicated tool (e.g. \cite{Simulator_z80}). ProB is a
general-purpose animator of formal specification, and addresses
computationally more difficult problems than a simulator. It is anyway
capable of animating, and also possibly model checking, a formal model
of the instruction set of the Z80, which was our initial motivation in
constructing that model. One may simulate step-by-step sequences of
Z80 instructions and visualize the visited states of ports, memory and
registers. It is a useful tool to debug by creating a trace of the
visited states, offering to analyse constants, functions and
expressions, and search possible violations.

The construction of the model with B generates a large number of proof
obligations and the verification is a time-consuming task. We describe
here some details on this effort and some solutions we have devised to
reduce the effort.  Recall that, in B, the generated proof obligations
allow to verify the data satisfy invariant properties, check important
system properties and that the expressions are well-defined
(WD)\footnote{An expression is called ``well-defined'' (or
  unambiguous) if its definition assigns it a unique interpretation or
  value.}.

@@

% Porque o modelo não foi verificado completamente ?
There is still a large number of open proof obligations, i.e. we have
not yet found their proof, or they are not valid and our model is not
correct. Each such proof obligation requires interaction with a proof
assistant. For a project of that size, a large effort needs to be
dedicated to find those proofs. Note that, once a new proof has been
found, the interactive prover can automatically apply it to other
proof obligations. It is often the case that the same proof script
discharges several proof obligations.

% ->Quanto tempo foi gasto nessa verificação ?  Aproximadamente quanto tempo de
% processamento elas reduziram ?
We had a previous experience with a similar effort, where we were able
to discharge all the proof obligations: that of the model of the Z80
instruction set designed without restricting ourselves to constructs
that may be animated with ProB, and without modeling external actions.
In that first experience, the verification took us about two
months. In that second project, we have been able to verify 3062
(e.g. roughly 80 \%) of 3631 proof obligations in one week.

In order to be able to verify such a large number of proof obligations
in such a short term, we employed two strategies: parallelism and
\emph{user pass}. AtelierB has an option to run any number of threads
in parallel and also to send proof obligations to other, possibly
remote, processes running AtelierB. With those options we had the
possibility to have eight concurrent instances of the prover. We thus
split the instruction set of the Z80 in eight sub-modules and checked
them concurrently (each instance of AtelierB may only verify one
operation at a time).

@@

The Z80 formal model provides many benefits, because of the
verification of generated proof obligations guarantees: the correct
use of data types, the developed security properties and the
well-defined of all the expressions.  Furthermore, the designer may
create new and specific security properties to adjust the verification
in accordance with the requirements of a specific
application. Moreover, this model complements the existing
documentation for users and assembly programmers.

But our fundamental goal is to use the model of the Z80 instruction
set to extend the scope of the B method up to the assembly
level~\cite{Dantas_SBMF08}. Indeed, we realized a case study where an
abstract B model had been refined up to the algorithmic model using
the classic development strategy. The algorithmic model was manually
compiled to a Z80 assembly program. We then constructed a B module
composed with the instructions of that program, and importing the
formal model of the Z80 instruction set. We then showed that this
assembly-level module was indeed a refinement of the original abstract
model.

@@

There exists several approaches to model hardware and instruction set
using B, not to mention other formal modelling approaches.  The paper
\cite{springerlink:Yuan2011} reports a method to specify, design and
construct sound and complete instruction set model by stepwise
refinement and formal proof using Event-B. It also discusses desirable
properties of an instruction set model. However, our work
\cite{Valerio_SBMF09} and \cite{Subotic2010} use the B method, that
seems quite appropriated to software development, because it has an
implementable language defined, called B0, a model-driven approach
\cite{Dantas_SBMF08} to develop software from the functional
specification level down to assembly, and tools to convert the models
to a programming language.

@@

In the future, we plan to improve the support for the development of
software with the B method from functional specification to assembly
level, using the Z80 model presented in this work. For instance, the
mechanic compilation from B algorithmic constructs to assembly
platform is also envisioned. Finally, another important activity is to
develop a formal model of the internal representation used in LLVM
Compiler~\cite{DBLP:conf/cgo/LattnerA04}, which would enable us to
integrate different compilation techniques into our B-based framework.
